An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-05-16 cal_demo/ms.tex
Fixing more terminology, going through sec 2 with fine toothed comb

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig1.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered 
electric fields are collected by the antenna elements then frequency transformed in the F-
engine. The calibration is applied, then the electric field measurements are gridded using the 
antenna voltage patterns. The imager performs a spatial 2D FFT on the gridded data before 
squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

@@

Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture distribution, $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. Although any antenna extends over a region of space, we include a nominal antenna 
position, $\mathbf{r}_a$, for multiple reasons. In the case where all antennas of an array are 
identical, a global aperture distribution can be adopted and the translation of the antennas 
distinguishes them. However, even if all antennas have unique aperture distributions, it can be 
convenient to retain the relative positions of antennas in the analysis. This allows for easy 
comparison with visibility-based analyses where the baseline vectors are used to describe 
sampling of the $uv$ plane. From a practical point of view, measurements of the antenna 
radiation pattern may not be performed in-situ, so it is convenient to describe the antenna 
aperture distribution relative to a point fixed within the aperture, along with a translational 
position.

@@

The next step is to grid the calibrated electric field measurements to a finites set of regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna aperture distributions as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
Mathematically we can express the gridding operation as
\begin{equation}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
This 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every sample, before any averaging of the data.

After gridding, the imaging portion of the MOFF performs a 2D spatial FFT to form 
instantaneous electric field images of the sky at fixed locations $\s_k$.
The Fourier dual coordinates to $\mathbf{r}$ are $(l,m)$, where $l=\sin\theta\cos\phi$,
$m=\nobreak\sin\theta\sin\phi$, and $(\theta,\phi)$ are the zenith 
and azimuthal angles. The full unit vector towards the sky is 
$\s=(l,m,\sqrt{1-l^2-m^2})$.
The sample spacing and resolution of $\s_k$ are given 
by the choices of $\mathbf{r}_j$ grid parameters, according to the usual FFT relations.
The pipeline up to this point can be summarized 
through the following equation.
\begin{align}
\mathcal{E}'(\s_k,t) & = \underbrace{\frac{1}{\Nant} \sum_j e^{2\pi i \mathbf{r}_j \cdot \s_k}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a W_a(\mathbf{r}_k - \ra)}_{\mathrm{Gridding}} \\
& \qquad \times \underbrace{ \frac{1}{g'_a} \left(g_a E_a^I(t)+n_a(t)\right)}_{\mathrm{Calibration}} \nonumber
\end{align}
The sum over antennas uses the antenna aperture distribution to grid the calibrated electric field measurements onto 
regular gridpoints, $\mathbf{r}_j$. The sum over $j$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums and use the DFT to approximate the Fourier relationship between
the antenna aperture distribution and the far-field radiation pattern.
\begin{align}\label{eq:epix}
\mathcal{E}(\s_k,t) & = \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_a E^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra} \nonumber \\
  & \qquad\times \sum_j W_a(\mathbf{r}_j-\ra)e^{2\pi i \s_k \cdot (\mathbf{r}_j-\ra)} \nonumber\\
& \approx \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_aE^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra}\mathcal{W}_a(\s_k)
\end{align}
In this form we see that the effect of gridding with the antenna aperture distribution is to attenuate 
the image by a factor of the far-field radiation response, $\mathcal{W}_a(\s)$.

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the complex conjugate of an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Cna}
\Kna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
the superscript $^*$ indicates complex conjugation,
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, 
and substituting the $g^{(n)}_b$ for $g'_b$ to indicate the specific loop number,
we find,
\begin{align}
\Kna & = \left<\left(g_a E_a^I(t)+n_a(t)\right) \frac{1}{\Nant} \right. \nonumber \\
& \qquad \left. \times \sum_b \frac{1}{g^{*(n)}_b}\left(g^*_bE_b^{I*}(t)+n^*_b(t)\right) e^{-2\pi i \spix \cdot \rb} \mathcal{W}^*_b(\spix)\right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left<\left(g_aE_a^I(t)+n_a(t)\right)\left(g^*_bE_b^{I*}+n^*_b(t)\right) \right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left(g_a g^*_b \left<E^I_a(t)E^{*I}_b(t)\right>_t + g_a\left<E_a^I(t)n^*_b(t)\right>_t \right. \nonumber \\
& \qquad \left. + g^*_b\left<n_a(t)E^{I*}_b(t)\right>_t + \left<n_a(t) n^*_b(t)\right>_t\right)  
%& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left(g_ag^*_bV^I_{ab}+n_{ab}\right)
\end{align}
where in the second step we group time-dependent terms, 
and in the third step we expanded the product within the time average.
Because the noise does not correlate with the ideal antenna measurements we can
drop the middle terms. We will also assume that the noise of antenna $a$ correlated with
the noise of antenna $b$ is also zero mean, unless $a=b$. By defining an ideal visibility as
$V_{ab}^I \equiv \left<E^I_a(t)E^{*I}_b(t)\right>_t$, we can simplify to
\begin{align}\label{eq:Kna}
\Kna & = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb}\nonumber \\
& \qquad \times \left( g_ag_b^* V_{ab}^I + \delta_{ab}\left<|n_a(t)|^2\right>_t\right)
\end{align}
where $\delta_{ab}$ is the Kronecker-delta function. To be clear, visibilities are never directly 
formed from the data. Rather it is convenient to express $\Kna$ in terms of the ideal 
visibilities, which we can model when solving for the gains below.

We will ultimately wish to model the right side of this equation to in order to solve for the
antenna gains. However, the self-correlated noise bias, $\left<|n_a(t)|^2\right>$, can often be 
significantly larger than the cross correlation power, and difficult to model precisely. It 
is therefore beneficial to subtract the self-correlation directly from $\Kna$. By properly
accounting for the various weighting and phasing terms, we can use self-correlation of the
uncalibrated antenna electric field measurements to remove the noise bias and $V^I_{aa}$
term from equation~\ref{eq:Kna}.
\begin{align}\label{eq:cna}
\Cna & \equiv \Kna - \frac{1}{\Nant g^{*(n)}_a}\mathcal{W}^*_a(\spix)e^{-2\pi i \spix \cdot \ra} \left<|E_a|^2\right>_t\nonumber \\
& = \frac{g_a}{\Nant} \sum_{b\ne a} \frac{g^*_b}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^I_{ab}
\end{align}

The next step is to use this antenna-pixel correlation to update our gain solution. We will model 
the right hand side of equation~\ref{eq:cna} by assuming our current gain estimates are 
approximately correct, $g^{(n)}_b\approx g_b$, and replacing the ideal visibilities with a set of 
model visibilities, $V^M_{ab}$. These model visibilities rely on knowledge of the sky, and should 
be precomputed before observing. We can then solve equation~\ref{eq:cna} for $g_a$ to achieve an 
updated estimate.
\begin{equation}\label{eq:cal_solution}
g^{(n+1)}_a = \Cna \Nant \left[ \sum_{b\ne a} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^M_{ab} \right]^{-1}
\end{equation}
This equation is our prescription for estimating the antenna gains of a direct imaging array. The 
real-time computation complexity scales as $\mathcal{O}(\Nant)$ as we form 
an antenna-pixel correlation, $\Cna$, for each 
antenna.

\textcolor{red}{pick up here}

We show schematically the process of calibrating a direct imaging correlator in 
figure~\ref{fig:schematic}. Computationally expensive steps that must be performed ``on-chip" 
are shown inside the gray box. The uncalibrated antenna signals are tapped out after the F-
engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is 
``on-chip''). The correlated values are then passed off-chip to estimate the gains using 
equation~\ref{eq:cal_solution}, and additional fitting if desired. The gains are then passed back 
to the correlator to update the calibration for subsequent integration intervals.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig2.pdf}
\caption{The general data flow of the MOFF correlator, with a feedback calibration loop. A pixel 
from the (unsquared) image is tapped out and correlated against the input antenna electric field 
signals to form $\Cna$ coefficients (equation~\ref{eq:Cna}). These coefficients are then 
shipped off-chip to apply equations~\ref{eq:cal_solution}~and~\ref{eq:damping}, and any other 
processing to arrive at an updated estimate of the gains, $g^{(n)}_a$. The gray box shows 
operations which must be done at high speed (before averaging in time), which white boxes 
show operations which can be performed ``off-chip".}
\label{fig:schematic}
\end{center}
\end{figure}

While testing we found equation~\ref{eq:cal_solution} resulted in oscillatory gain solutions as it 
was iterated, as is often the case in iterative minimization methods. To mitigate this we 
introduce a damping factor, $0 \leq \damp <1$, which is used to attenuate the gain update, 
effectively giving the solutions memory of previous iterations.
\begin{equation}\label{eq:damping}
g^{(n+1)}_a = (1-\damp) g'^{(n+1)}_a + \damp g^{(n)}_a
\end{equation}
We found that while equation~\ref{eq:cal_solution} does indeed converge on good solutions, 
the process is made faster by tuning the damping factor. While $g'^{(n+1)}_a$ is the best 
estimate of the gains after the $n^{\textrm{th}}$ iteration, the damped version, $g^{(n+1)}_a$, is 
actually used in the iterative calibration loop. Once the loop converges the damped version is 
essentially a weighted average over the past several iterations, giving it a longer effective 
integration time.

An important feature to note is that, like the MOFF-generated images themselves, 
equation~\ref{eq:cal_solution} includes the antenna auto-correlations (the sum is over 
\emph{all} $b$, not excluding $a$). It can be difficult to perfectly model the noise bias from 
auto-correlations, which can often times be far brighter than the visibilities themselves. It can 
therefore be beneficial to subtract this term directly from \Cna, and exclude the $b=a$ term in 
the sum.
\begin{equation}
\Cna \rightarrow \Cna - \frac{1}{\Nant\, g^{*(n)}_a} \beamtheta^*_a(\spix) e^{-2\pi i \spix \cdot \ra} \left<|\Er{a}|^2\right>_t
\end{equation}
We have absorbed $g_a^T$ into the electric field so that the correlation is now simply the 
autocorrelation of the uncalibrated antenna measurement. This requires generating these 
correlations, which again only scale as $\mathcal{O}(\Nant)$, and are generally useful for array 
diagnostics.

We conclude this section by connecting our calibration expression to that found in a visibility 
framework. In the limit of a single bright calibrating source at phase center, we can greatly 
simplify equations~\ref{eq:cna} and~\ref{eq:cal_solution}. We will assume the beams are 
normalized such that $\beamtheta(0)=1$. We can further drop the exponential phase terms 
because $\spix=0$. We then absorb the true gains into the true visibilities in 
equation~\ref{eq:cna} to express as a sum of measured, uncalibrated, visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_b \frac{1}{g^{*(n)}_b} \V_{ab}
\end{equation}

We next plug this expression into equation~\ref{eq:cal_solution} to find our simplified 
calibration solution for a single bright point source. Because our sky is a single bright point 
source, the model visibilities are simply the flux of the source, $S_{\mathrm{src}}$.
\begin{equation}
g'^{(n+1)}_a \rightarrow \frac{\sum_b  \V_{ab}/g^{*(n)}_b}{\Nant S_{\mathrm{src}}}
\end{equation}
This is simply a gain-weighted sum of the measured visibilities over the flux of the source, 
which is indeed the limiting result from a visibility approach, for example seen in \citealt{mit08}. 
The ability to recover the equivalent expression despite not actually forming the visibilities is a 
result of the fact that only sums over visibilities come into the FX solution, as was described in 
\citealt{mor11}. We have confirmed the limiting case equivalence here, and will explore the 
more general case in more detail in \S~\ref{sec:noise}.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig1.pdf}
\caption{Data flow for the MOFF algorithm, reproduced from \citealt{thy15c}. The time ordered 
electric field measurements are collected by the antenna elements then frequency transformed in the 
F-engine. The calibration is applied, then the electric field measurements are gridded using the 
antenna illumination patterns. The imager performs a spatial 2D FFT on the gridded data before 
squaring and averaging. The accumulated images are then written to disk.}
\label{fig:moff_flow}
\end{center}
\end{figure}

@@

Next we turn to the measurement performed by an antenna with far-field radiation pattern 
$\mathcal{W}_a(\hat{\mathbf{s}})$, where the subscript $a$ denotes a specific antenna. We 
adopt the common convention of normalizing the radiation pattern to have response of 1 at the 
center of the main lobe \citep[e.g.][]{nap99}. Consequently, the translation between the terminal 
voltages to electric field measurements has been absorbed in our definition of the radiation 
pattern. The ideal antenna electric field measurement can be expressed as 
\begin{equation}
E^I_a(f,t) = \int \mathcal{W}_a(\hat{\mathbf{s}})\mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega. 
\end{equation}
Equivalently we can express the effect of the antenna radiation pattern in the plane of the 
aperture.
\begin{equation}
E^I_a(f,t) = \int_{\mathrm{aperture}}W_a(\mathbf{r}-\mathbf{r}_a)E_a(\mathbf{r},f,t)\,d^2\mathbf{r}
\end{equation}
Here we used the Fourier relationship between the antenna's far-field radiation pattern and the 
aperture illumination pattern (sometimes referred to as the aperture distribution), $W_a(\mathbf{r}-\mathbf{r}_a)=\int  \mathcal{W}_a(\hat{\mathbf{s}}) \, d
\Omega$. Although any antenna extends over a region of space, we include a nominal antenna 
position, $\mathbf{r}_a$, for multiple reasons. In the case where all antennas of an array are 
identical, a global aperture distribution can be adopted and the translation of the antennas 
distinguishes them. However, even if all antennas have unique aperture distributions, it can be 
convenient to retain the relative positions of antennas in the analysis. This allows for easy 
comparison with visibility-based analyses where the baseline vectors are used to describe 
sampling of the $uv$ plane. From a practical point of view, measurements of the antenna 
radiation pattern may not be performed in-situ, so it is convenient to describe the antenna 
aperture distribution relative to a point fixed within the aperture, along with a translational 
position.

@@

The next step is to grid the calibrated electric field measurements to a finite set of regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna aperture distributions as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
Mathematically we can express the gridding operation as
\begin{equation}\label{eq:gridding}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
Formally each pixel is calculated by summing over all antennas and their illumination patterns.
However the illumination patterns are typically compact and non-overlapping, so most pixels
will only contain contribution from at most one antenna, allowing equation~\ref{eq:gridding}
to be calculated through a sparse matrix multiplication.
This gridding 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every time sample, before any averaging of the data.

After gridding, the imaging portion of the MOFF performs a 2D spatial FFT to form 
instantaneous electric field images of the sky at fixed locations $\s_k$.
The Fourier dual coordinates to $\mathbf{r}$ are $(l,m)$, where $l=\sin\theta\cos\phi$,
$m=\nobreak\sin\theta\sin\phi$, and $(\theta,\phi)$ are the zenith 
and azimuthal angles. The full unit vector towards the sky is 
$\s=(l,m,\sqrt{1-l^2-m^2})$.
The sample spacing and resolution of $\s_k$ are given 
by the choices of $\mathbf{r}_j$ grid parameters, according to the usual FFT relations. 
Fine $\mathbf{r}_j$ grid spacing can result in non-physical angular pixels ($l^2+m^2>1$),
which should be discarded.

The pipeline up to this point can be summarized 
through the following equation.
\begin{align}
\mathcal{E}'(\s_k,t) & = \underbrace{\frac{1}{\Nant} \sum_j e^{2\pi i \mathbf{r}_j \cdot \s_k}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a W_a(\mathbf{r}_j - \ra)}_{\mathrm{Gridding}} \\
& \qquad \times \underbrace{ \frac{1}{g'_a} \left(g_a E_a^I(t)+n_a(t)\right)}_{\mathrm{Calibration}} \nonumber
\end{align}
The sum over antennas uses the antenna illumination pattern to grid the calibrated electric field measurements onto 
regular gridpoints, $\mathbf{r}_j$. The sum over $j$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums and use the DFT to approximate the Fourier relationship between
the antenna illumination pattern and the far-field radiation pattern.
\begin{align}\label{eq:epix}
\mathcal{E}(\s_k,t) & = \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_a E^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra} \nonumber \\
  & \qquad\times \sum_j W_a(\mathbf{r}_j-\ra)e^{2\pi i \s_k \cdot (\mathbf{r}_j-\ra)} \nonumber\\
& \approx \frac{1}{\Nant} \sum_a \frac{1}{g'_a}\left(g_aE^I_a(t)+n_a(t)\right) e^{2\pi i \s_k \cdot \ra}\mathcal{W}_a(\s_k)
\end{align}
In this form we see that the effect of gridding with the antenna illumination pattern is to attenuate 
the image by a factor of the far-field radiation response, $\mathcal{W}_a(\s)$.

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the complex conjugate of an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Kna_def}
\Kna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
the superscript $^*$ indicates complex conjugation,
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

Plugging equation \ref{eq:epix} into equation \ref{eq:Kna_def}, 
and substituting the $g^{(n)}_b$ for $g'_b$ to indicate the specific loop number,
we find,
\begin{align}
\Kna & = \left<\left(g_a E_a^I(t)+n_a(t)\right) \frac{1}{\Nant} \right. \nonumber \\
& \qquad \left. \times \sum_b \frac{1}{g^{*(n)}_b}\left(g^*_bE_b^{I*}(t)+n^*_b(t)\right) e^{-2\pi i \spix \cdot \rb} \mathcal{W}^*_b(\spix)\right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left<\left(g_aE_a^I(t)+n_a(t)\right)\left(g^*_bE_b^{I*}+n^*_b(t)\right) \right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left(g_a g^*_b \left<E^I_a(t)E^{*I}_b(t)\right>_t + g_a\left<E_a^I(t)n^*_b(t)\right>_t \right. \nonumber \\
& \qquad \left. + g^*_b\left<n_a(t)E^{I*}_b(t)\right>_t + \left<n_a(t) n^*_b(t)\right>_t\right)  
%& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left(g_ag^*_bV^I_{ab}+n_{ab}\right)
\end{align}
where in the second step we group time-dependent terms, 
and in the third step we expanded the product within the time average.
Because the noise does not correlate with the ideal antenna measurements we can
drop the middle terms. We will also assume that the cross-correlation noise ($a\ne b$) 
is zero mean. Then by defining an ideal visibility as
$V_{ab}^I \equiv \left<E^I_a(t)E^{*I}_b(t)\right>_t$, we can simplify to
\begin{align}\label{eq:Kna}
\Kna & = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb}\nonumber \\
& \qquad \times \left( g_ag_b^* V_{ab}^I + \delta_{ab}\left<|n_a(t)|^2\right>_t\right)
\end{align}
where $\delta_{ab}$ is the Kronecker-delta function. To be clear, visibilities are never directly 
formed from the data. Rather it is convenient to express $\Kna$ in terms of the ideal 
visibilities, which we can model when solving for the gains below.

We will ultimately wish to model the right side of equation~\ref{eq:Kna} in order to solve for the
antenna gains. However, the self-correlated noise, $\left<|n_a(t)|^2\right>$, can often be 
significantly larger than the cross correlation power, and difficult to model precisely. It 
is therefore beneficial to subtract the self-correlation directly from $\Kna$. By properly
accounting for the various weighting and phasing terms, we can use self-correlations of the
uncalibrated antenna electric field measurements to remove the corresponding
terms from equation~\ref{eq:Kna}.
\begin{align}\label{eq:Cna}
\Cna & \equiv \Kna - \frac{1}{\Nant g^{*(n)}_a}\mathcal{W}^*_a(\spix)e^{-2\pi i \spix \cdot \ra} \left<|E_a|^2\right>_t\nonumber \\
& = \frac{g_a}{\Nant} \sum_{b\ne a} \frac{g^*_b}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^I_{ab}
\end{align}

The next step is to use this antenna-pixel correlation to update our gain solution. We will model 
the right hand side of equation~\ref{eq:Cna} by assuming our current gain estimates are 
approximately correct, $g^{(n)}_b\approx g_b$, and replacing the ideal visibilities with a set of 
model visibilities, $V^M_{ab}$. These model visibilities rely on knowledge of the sky, and should 
be precomputed before observing. We can then solve equation~\ref{eq:Cna} for $g_a$ to achieve an 
updated estimate.
\begin{equation}\label{eq:cal_solution}
%g^{(n+1)}_a = \Cna \Nant \left[ \sum_{b\ne a} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^M_{ab} \right]^{-1}
g^{(n+1)}_a = \frac{\Cna \Nant }{ \sum_{b\ne a} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^M_{ab} }
\end{equation}
This equation is our prescription for estimating the antenna gains of a direct imaging array. The 
real-time computation complexity scales as $\mathcal{O}(\Nant)$ as we form 
an antenna-pixel correlation, $\Cna$, for each 
antenna.

We show schematically the process of calibrating a direct imaging correlator in 
figure~\ref{fig:schematic}. Computationally expensive steps that must be performed in near real-time 
are shown inside the gray box. The uncalibrated antenna signals are tapped out after the 
F-engine and correlated against the output image pixel of interest (i.e., equation~\ref{eq:Cna} is 
computed in near real-time). The correlated values are then used to estimate the gains
using equation~\ref{eq:cal_solution}, and additional fitting if desired. This step does not have
the speed requirement of the steps in the gray box because the correlations have been 
averaged in time.
The gains are then passed back 
to the correlator to update the calibration for subsequent integration intervals.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig2.pdf}
\caption{The general data flow of the MOFF correlator, with a feedback calibration loop. A pixel 
from the (unsquared) image is tapped out and correlated against the input antenna electric field 
measurements to form $\Cna$ coefficients (equation~\ref{eq:Cna}). These coefficients are then 
used to update the gain estimates using equations~\ref{eq:cal_solution}~and~\ref{eq:damping},
and any additional fitting.
The gray box shows operations which must be done at high speed (before averaging in time),
while the white boxes show operations which can be performed at lower speeds because
they involve time averages quantities.
}
\label{fig:schematic}
\end{center}
\end{figure}

While testing we found equation~\ref{eq:cal_solution} resulted in oscillatory gain solutions over time, 
as is often the case in iterative methods. To mitigate this we 
introduce a damping factor, $0 \leq \damp <1$, which is used to attenuate the gain update, 
effectively giving the solutions memory of previous iterations.
\begin{equation}\label{eq:damping}
%g^{(n+1)}_a = (1-\damp) g'^{(n+1)}_a + \damp g^{(n)}_a
g^{(n+1)}_a = \frac{(1-\damp)\Cna \Nant }{ \sum_{b\ne a} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^M_{ab} }+ \damp g^{(n)}_a
\end{equation}
We found that while equation~\ref{eq:cal_solution} does indeed converge on good solutions, 
the process is made faster by tuning the damping factor. 
 Once the loop converges the damped version is 
essentially a weighted average over the past several iterations, giving it a longer effective 
integration time.

We conclude this section by connecting our calibration expression to that found in a visibility 
framework. In the limit of a single bright calibrating source at center, we can greatly 
simplify equations~\ref{eq:Cna} and~\ref{eq:cal_solution}. We will assume the radiation patterns are 
normalized such that $\mathcal{W}(0)=1$. We can further drop the exponential phase terms 
because $\spix$ is perpendicular to our planar antenna array. We then absorb the true gains into the true visibilities in 
equation~\ref{eq:Cna} to express as a sum of measured, uncalibrated, visibilities.
\begin{equation}
\mathcal{C}^{(n)}_{a,0} \rightarrow \frac{1}{\Nant}\sum_b \frac{1}{g^{*(n)}_b} \V_{ab}
\end{equation}

We next plug this expression into equation~\ref{eq:cal_solution} to find our simplified 
calibration solution for a single bright point source. Because our sky is a single bright point 
source, the model visibilities are simply the flux of the source, $S_{\mathrm{src}}$.
\begin{equation}
g^{(n+1)}_a \rightarrow \frac{\sum_b  V_{ab}/g^{*(n)}_b}{\Nant S_{\mathrm{src}}}
\end{equation}
This is simply a gain-weighted sum of the measured visibilities over the flux of the source, 
which is indeed the limiting result from a visibility approach, for example seen in \citealt{mit08}. 
The ability to recover the equivalent expression despite not actually forming the visibilities is a 
result of the fact that only sums over visibilities come into the visibility-based solution, as was described in 
\citealt{mor11}. We have confirmed the limiting case equivalence here, and will explore the 
more general case in more detail in \S~\ref{sec:noise}.
